Model_additionalAlts

### 2015-12-01 models/Model_additionalAlts.Rmd
slight modifications and correct output for fitted literal listener

# RSA functionality
```{r}
# speaker informativity
# ---------------------
speaker.inf = function(d, alpha, cost = 0) {
  exp(alpha*(log(d) - cost))
}
# speaker likelihood
# ------------------
speaker.lhd = function(rating, degree, m, alpha) {
  numerator = speaker.inf(m[rating, degree], alpha)
  normalize = sum(sapply(m[rating, ], function(i) {speaker.inf(i, alpha)}))
  return(numerator / normalize)
}
# non-normalized posterior
# -----------------------
nn.post = function(rating, degree, m, alpha, useprior) {
  prior = priors[rating, "prior.p"]
  return(speaker.lhd(rating, degree, m, alpha) * prior)
}
# normalized posterior
# --------------------
norm.post = function(rating, degree, m, alpha, useprior) {
  nn = nn.post(rating, degree, m, alpha, useprior)
  normalize = sum(unlist(sapply(seq(1, 5), function(i){nn.post(i, degree, m, alpha, useprior)})))
  return(nn / normalize)
}
```

# Run model functionality
```{r}
# run.partial()
# ------------
# Run RSA with model1 (entailment) and model2 (entailment + generic)
run.partial = function(d, alpha=1, useprior=F, usenone=F, normalize=F) {
  if (normalize) {
    mat = d %>%
      select(stars, degree, speaker.p) %>%
      spread(degree, speaker.p) %>%
      mutate(hi = hi / sum(hi),
             low = low / sum(low)) %>%
      select(hi, low)
  } else {
    mat = d %>%
      select(stars, degree, speaker.p) %>%
      spread(degree, speaker.p) %>%
      select(hi, low)
  }

@@

## Tune hyperparams
```{r}
# tune.alhpa()
# ------------
# d        --> data
# alphas   --> range of alphas to test
# type     --> full or partial model
# useprior --> use uniform prior
# usenone  --> use gener None
tune.alpha = function(d, alphas = seq(from=1, to=10),
                      type="partial", useprior = T,
                      usenone=F, compare.data=NULL, addMid = F, normalize=F) {
  # Tune best alphas
  fit = sapply(alphas, FUN=function(n) {
    if (type == "partial") {
      md = d %>%
        do(run.partial(., alpha=n, useprior=useprior, usenone=usenone, normalize=normalize))
    } else {
      md = d %>%
        do(run.full(., alpha=n, useprior=useprior, usenone=usenone, addMid=addMid, normalize=normalize))
    }
    # Fit to e6 data
    if (!is.null(compare.data)) {
      matched.items = which((md[, "scale"] != "some_all" &
                               (md[, "degree"] == "hi2" | md[, "degree"] == "hi1")) |
                              (md[, "scale"] == "some_all" &
                                 (md[, "degree"] == "hi1" | md[, "degree"] == "low1")))
      md = md[matched.items, ]
      md$degree = ifelse(md$degree == "hi1", "hi", "low")
      stopifnot("listener.p" %in% colnames(compare.data))
      md$listener.p = compare.data$listener.p
    }

@@

### Data for full model - alternatives
```{r}
speaker = read.csv("~/Desktop/Projects/scalar_implicature/models/model_data/L0_e10a.csv")
data.full.extras = left_join(speaker, listener) %>%
  left_join(priors) %>%
  rowwise %>%
  select(scale, degree, stars, speaker.p, listener.p, prior.p) %>%
  mutate(listener.p = ifelse(is.na(listener.p), 0, listener.p)) %>%
  group_by(scale)
```

## Match items between studies (full and partial)
```{r}
matched.items = which((data.full[, "scale"] != "some_all" &
        (data.full[, "degree"] == "hi2" | data.full[, "degree"] == "hi1")) |
       (data.full[, "scale"] == "some_all" &
          (data.full[, "degree"] == "hi1" | data.full[, "degree"] == "low1")))
matched.items.extras = which((data.full.extras[, "scale"] != "some_all" &
        (data.full.extras[, "degree"] == "hi2" | data.full.extras[, "degree"] == "hi1")) |
       (data.full.extras[, "scale"] == "some_all" &
          (data.full.extras[, "degree"] == "hi1" | data.full.extras[, "degree"] == "low1")))
```

# Fit extra scalar literal listener distributions
## modified gradient descent / MLE fn()
```{r}
# normalize()
# -----------
# normalize a generic vector
normalize = function(w) { return (w / sum(w))}

@@

# Fit extra scalar literal listener distributions
```{r}
# Normalize in RSA?
norm = T
# Our scalars
scalars = c("good_excellent",
            "liked_loved",
            "palatable_delicious",
            "memorable_unforgettable",
            "some_all")
fitted.distr = data.frame(matrix(rep(NA, 25), ncol=5))
colnames(fitted.distr) = scalars

# Fit each of our scalars / track fitting
track.data = data.frame(matrix(ncol=6))
for (i in 1:length(scalars)) {
  # record fitting
  data.record = data.frame(matrix(rep(NA, 5), ncol=5))
  # output is optimal distr [[1]] and [[2]] record of fitting
  out = fit_distr(scalars[i], numIters=20, eta=0.01,
                  verbose=F, normalize=norm, alpha=4, data.record=data.record)

@@

```{r}
# Format track data for display
track.data = track.data[-1, ]
track.data$step = rep(seq(1, 20), 5)
track.data = track.data %>% 
  gather(stars, value, X1:X5)
```

## Run distribution fitting
```{r}
for (scale in scalars) {
  data.full.extras[data.full.extras$scale == scale &
                     data.full.extras$degree == "mid", ]$speaker.p =
    normalize(fitted.distr[ , scale])
}
m = data.full.extras %>%
  do(run.full(., alpha = 4, addMid=T, normalize=norm))
cor(m[matched.items.extras, ]$listener.p, m[matched.items.extras, ]$pred)
```

## Plot distr
```{r}
# Normalize speaker.p for plotting
for (scale in scalars) {
  degrees = unique(data.full.extras$degree)
  for (degree in degrees) {
    data.full.extras[data.full.extras$scale == scale &
                       data.full.extras$degree == degree, ]$speaker.p = 
        normalize(data.full.extras[data.full.extras$scale == scale &
                                     data.full.extras$degree == degree, ]$speaker.p)
  }
}

# Differentiate tuned 'mid' scalar
data.full.extras$type = ifelse(data.full.extras$degree == "mid", "red", "blue")
qplot(x = stars, y = speaker.p, colour = type,
      geom = "line", stat = "identity", position = "dodge", 
      main = "Literal listener semantics\nEmpirical estimates and tuned extra 'mid'",
      ylab = "P_{L0}(Word | Rating) - Normalized",
      data = data.full.extras) + 
  facet_wrap(~scale + degree)

# Tracking of gradient descent tuning
ggplot(data=track.data, aes(x=step, y=value, col=stars)) +
  geom_line() +
  facet_wrap(~X6)
```

>>>>

# RSA functionality
```{r RSA_function}
# speaker informativity
# ---------------------
speaker.inf = function(d, alpha, cost = 0) {
  exp(alpha*(log(d) - cost))
}
# speaker likelihood
# ------------------
speaker.lhd = function(rating, degree, m, alpha) {
  numerator = speaker.inf(m[rating, degree], alpha)
  normalize = sum(sapply(m[rating, ], function(i) {speaker.inf(i, alpha)}))
  return(numerator / normalize)
}
# non-normalized posterior
# -----------------------
nn.post = function(rating, degree, m, alpha, useprior) {
  prior = priors[rating, "prior.p"]
  return(speaker.lhd(rating, degree, m, alpha) * prior)
}
# normalized posterior
# --------------------
norm.post = function(rating, degree, m, alpha, useprior) {
  nn = nn.post(rating, degree, m, alpha, useprior)
  normalize = sum(unlist(sapply(seq(1, 5), function(i){nn.post(i, degree, m, alpha, useprior)})))
  return(nn / normalize)
}
```

# Run model functionality
```{r run_functions}
# run.partial()
# ------------
# Run RSA with model1 (entailment) and model2 (entailment + generic)
run.partial = function(d, alpha=1, useprior=F, usenone=F, normalize=F) {
  if (normalize) {
    mat = d %>%
      select(stars, degree, speaker.p) %>%
      spread(degree, speaker.p) %>%
      mutate(hi = hi / sum(hi),
             low = low / sum(low)) %>%
      select(hi, low)
  } else {
    mat = d %>%
      select(stars, degree, speaker.p) %>%
      spread(degree, speaker.p) %>%
      select(hi, low)
  }

@@

## Tune hyperparams
```{r alpha_tuning}
# tune.alhpa()
# ------------
# d        --> data
# alphas   --> range of alphas to test
# type     --> full or partial model
# useprior --> use uniform prior
# usenone  --> use gener None
tune.alpha = function(d, alphas = seq(from=1, to=10),
                      type="partial", useprior = T,
                      usenone=F, compare.data=NULL, addMid = F, normalize=F) {
  # Tune best alphas
  fit = sapply(alphas, FUN=function(n) {
    if (type == "partial") {
      md = d %>%
        do(run.partial(., alpha=n, useprior=useprior, usenone=usenone, normalize=normalize))
    } else {
      md = d %>%
        do(run.full(., alpha=n, useprior=useprior, usenone=usenone, addMid=addMid, normalize=normalize))
    }
    # Fit to e6 data
    if (!is.null(compare.data)) {
      matched.items = which((md[, "scale"] != "some_all" &
                               (md[, "degree"] == "hi2" | md[, "degree"] == "hi1")) |
                              (md[, "scale"] == "some_all" &
                                 (md[, "degree"] == "hi1" | md[, "degree"] == "low1")))
      md = md[matched.items, ]
      md$degree = ifelse(md$degree == "hi1", "hi", "low")
      stopifnot("listener.p" %in% colnames(compare.data))
      md$listener.p = compare.data$listener.p
    }

@@

### Data for full model - alternatives
```{r full_model_data + alternatives, message=FALSE, warning=FALSE}
speaker = read.csv("~/Desktop/Projects/scalar_implicature/models/model_data/L0_e10a.csv")
data.full.extras = left_join(speaker, listener) %>%
  left_join(priors) %>%
  rowwise %>%
  select(scale, degree, stars, speaker.p, listener.p, prior.p) %>%
  mutate(listener.p = ifelse(is.na(listener.p), 0, listener.p)) %>%
  group_by(scale)
```

## Match items between studies (full and partial)
```{r match_items_between_studies}
matched.items = which((data.full[, "scale"] != "some_all" &
        (data.full[, "degree"] == "hi2" | data.full[, "degree"] == "hi1")) |
       (data.full[, "scale"] == "some_all" &
          (data.full[, "degree"] == "hi1" | data.full[, "degree"] == "low1")))
matched.items.extras = which((data.full.extras[, "scale"] != "some_all" &
        (data.full.extras[, "degree"] == "hi2" | data.full.extras[, "degree"] == "hi1")) |
       (data.full.extras[, "scale"] == "some_all" &
          (data.full.extras[, "degree"] == "hi1" | data.full.extras[, "degree"] == "low1")))
```

# Fit extra scalar literal listener distributions
## modified gradient descent / MLE fn()
```{r Fit extra scalars functions, warning=FALSE}
# normalize()
# -----------
# normalize a generic vector
normalize = function(w) { return (w / sum(w))}

@@

# Fit extra scalar literal listener distributions
```{r Run fit extra scalars}
# Set parameters for tunin
# Normalize in RSA?
norm = T
eta = 0.01
alpha = 4
numIters = 100

# Our scalars
scalars = c("good_excellent",
            "liked_loved",
            "palatable_delicious",
            "memorable_unforgettable",
            "some_all")
fitted.distr = data.frame(matrix(rep(NA, 25), ncol=5))
colnames(fitted.distr) = scalars

# Fit each of our scalars / track fitting
track.data = data.frame(matrix(ncol=6))
for (i in 1:length(scalars)) {
  # record fitting
  data.record = data.frame(matrix(rep(NA, 5), ncol=5))
  # output is optimal distr [[1]] and [[2]] record of fitting
  out = fit_distr(scalars[i], numIters=numIters, eta=eta,
                  verbose=F, normalize=norm, alpha=alpha, data.record=data.record)

@@

```{r format tracking data}
# Format track data for display
track.data = track.data[-1, ]
track.data$step = rep(seq(1, numIters), 5)
track.data = track.data %>% 
  gather(stars, value, X1:X5)
track.data$stars = ifelse(track.data$stars == "X1", "1",
                          ifelse(track.data$stars == "X2", "2",
                                 ifelse(track.data$stars == "X3", "3",
                                        ifelse(track.data$stars == "X4", "4", 5))))
colnames(track.data)[1] = "scale"
```

## Run distribution fitting
```{r Run-data}
for (scale in scalars) {
  data.full.extras[data.full.extras$scale == scale &
                     data.full.extras$degree == "mid", ]$speaker.p =
    normalize(fitted.distr[ , scale])
}
m = data.full.extras %>%
  do(run.full(., alpha = alpha, addMid=T, normalize=norm))
cor(m[matched.items.extras, ]$listener.p, m[matched.items.extras, ]$pred)
```

## Plot distr
```{r Plots}
# Normalize speaker.p for plotting
for (scale in scalars) {
  degrees = unique(data.full.extras$degree)
  for (degree in degrees) {
    data.full.extras[data.full.extras$scale == scale &
                       data.full.extras$degree == degree, ]$speaker.p = 
        normalize(data.full.extras[data.full.extras$scale == scale &
                                     data.full.extras$degree == degree, ]$speaker.p)
  }
}

# Differentiate tuned 'mid' scalar
plot.data = data.full.extras
plot.data$type = ifelse(plot.data$degree == "mid", "red", "blue")
qplot(x = stars, y = speaker.p, colour = type,
      geom = "line", stat = "identity", position = "dodge", 
      main = "Literal listener semantics\nEmpirical estimates and tuned extra 'mid'",
      ylab = "P_{L0}(Word | Rating) - Normalized",
      data = plot.data) + 
  facet_wrap(~scale + degree)

# Tracking of gradient descent tuning
ggplot(data=track.data, aes(x=step, y=value, col=stars)) +
  geom_line() +
  facet_wrap(~scale) +
  ylab("Normalized p(m | u)") +
  xlab("Iterations") +
  ggtitle("Fitted alternative distr fitting")
```

## Output fitted literal listener to .csv
```{r}
data.out = data.full.extras %>%
  select(degree, stars, speaker.p) %>%
  mutate(cil = NA,
         cih = NA,
         cnt.judgment = NA)
data.out = data.out[, c(1, 2, 3, 5, 6, 7, 4)]
# write.csv(data.out,
#           "~/Desktop/Projects/scalar_implicature/models/model_data/L0_e10fittedAlts.csv")
```
